---
title: "HCA Data Discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HCA Data Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE
)
```

```{r setup, include = FALSE}
library(SingleCellExperiment)
library(LoomExperiment)
library(dplyr)
library(HCAMatrixBrowser)
```

Objectives:

- Explore the HCA data portal via web browser.
- Download and import an example data matrix.
- Illustrate essential operations on single cell data in R / Bioconductor.

# The HCA Data Portal

The HCA [data portal][HCADataPortal]

- https://data.humancellatlas.org/explore

- An example (chosen arbitrarily!): [Assessing the relevance of
  organoids to model inter-individual variation][eg1].

[eg1]: https://data.humancellatlas.org/explore/projects/005d611a-14d5-4fbf-846e-571a1f874f70

Different ways to export data.

- csv, mtx, loom.
- Loom format is richest, with abundant metadata about cells and genes.

# Data retrieval and import

Retrieve the data

```{r}
url <- paste0(
    "https://data.humancellatlas.org/project-assets/project-matrices/",
    "005d611a-14d5-4fbf-846e-571a1f874f70.homo_sapiens.loom"
)
path <- BiocFileCache::bfcrpath(rnames = url)
```

- Use [BiocFileCache][] to retrieve once, for fast easy access the next time.

Load the data using [LoomExperiment][]

```{r}
library(LoomExperiment)
loom <- import(path)
loom
```

- A `LoomExperiment` is much like a `SummarizedExperiment`.
- Transforms internal file-based representation of native `loom`
  format to familiar R / Bioconductor structures.

# Exploring single cell data

Easy to sample and explore

```{r}
x0 <- as.matrix(assay(loom)[, 1:1000])
library_size <- colSums(x0)
non_zero_genes <- colSums(x0 > 0)
```

- Create a subset of data and represent as in-memory matrix.
- 'Library size' as sum of counts in each column.
- How many genes with non-zero expression in each cell? Determine
  which counts are greater than 0 (`TRUE` or `FALSE`), sum the number
  of `TRUE` in each column.

```{r}
plot(library_size ~ non_zero_genes, log = "xy")
abline(v = 500)
```

- Bimodal distribution of cells make us appreciate need for quality
  assessment, etc.

Extensive row and column data annotations

```{r}
head(rowData(loom), 3)
```

- Annotations on genes from the HCA pipeline used in pre-processing.

```{r}
library(dplyr)
colData(loom) %>%
    as_tibble()
```

- Very rich data cell-level data summarized by the HCA
- Can be linked to even more annotation information!

Easily coerce to standard `SingleCellExperiment`...

```{r}
sce <- as(loom, "SingleCellExperiment")
names(assays(sce)) <- "counts"
sce
```

- ...ready for down-stream exploration and analysis.

# Writing .loom files

Use `export()` to create a loom file for sharing with others.

```{r, eval = FALSE}
my_loom <- tempfile(fileext =".loom")
export(as(sce, "LoomExperiment"), my_loom)
```

[BiocFileCache]: https://bioconductor.org/packages/BiocFileCache
[LoomExperiment]: https://bioconductor.org/packages/LoomExperiment

# Summary

- Pre-computed data matrices are easy to discover from [HCA Data
  Portal][HCADataPortal]
- `.loom` format provides very rich row (gene) and column (cell)
  annotations.
- [LoomExperiment][] imports data into R in a memory-efficient manner
- It is easy to integrate LoomExperiment objects into the rich R /
  Bioconductor infrastructure for downstream analysis.
  
## Session info

```{r}
sessionInfo()
```

[HCADataPortal]: https://data.humancellatlas.org/explore
