---
title: "Programmatic exploration of HCA data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programmatic exploration of HCA data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE
)
```

```{r setup, include = FALSE}
library(HCAMatrixBrowser)
```

Objectives:

- Explore HCA matrix data programmatically
- Use the [HCAMatrixBrowser][] 'high level' interface to filter,
  create, and retrieve a [LoomExperiment][] with a custom count matrix.
- Introduce [HCAMatrixBrowser][] 'low level' operations for matrix retrieval.

[HCAMatrixBrowser]: https://bioconductor.org/packages/HCAMatrixBrowser
[LoomExperiment]: https://bioconductor.org/packages/LoomExperiment

# The HCAMatrixBrowser package

```{r}
library(HCAMatrixBrowser)
print(citation("HCAMatrixBrowser"), style = "text")
```

Built on the [HCA Matrix Service][] RESTful API.

```{r}
hca <- HCAMatrix()
```

- Create an object that references the service.

[HCA Matrix Service]: https://matrix.data.humancellatlas.org/

# 'High-level' use

Discover available filters.

```{r}
filters <- available_filters(hca)
length(filters)
grep("project", filters, value = TRUE)
```

- Filters available on 43 different facets
- The `project.project_core.project_title` filter corresponds to
  titles on the matrix browser web page.

Select a particular experiment

```{r}
title <- "Assessing the relevance of organoids to model inter-individual variation"
hca1 <- filter(
    hca,
    project.project_core.project_title == title
)
filters(hca1) %>%
    str()
```

Combine filters

```{r}
grep("detected", filters, value = TRUE)
filter_detail(hca, "genes_detected") %>%
    str()
```

- `genes_detected` can be used to filter only cells with more than 500
  genes observed.
- Combine with project selection.

```
hca1 <- filter(
    hca,
    project.project_core.project_title == title &
    genes_detected >= 500
)
filters(hca1)
```

# Retrieve the custom selection of cells

```
> loom <- loadHCAMatrix(hca1, format = "loom")
HCAMatrixBrowser cache directory set to:
    /Users/ma38727/Library/Caches/org.R-project.R/R/HCAMatrixBrowser
  (\) waiting for query completion.......... 00:00:02
  (-) waiting for query completion.......... 00:00:06
...
  (/) waiting for query completion.......... 00:08:13
  |======================================================================| 100%
```

- Matrix assembled by the service (remotely) 'on the fly'

```
> loom
class: LoomExperiment 
dim: 58347 43861 
metadata(0):
assays(1): matrix
rownames: NULL
rowData names(9): Accession Gene ... genus_species isgene
colnames(43861): 0006d1a2669431b298d08f010da6480b
  000f429f8f0c461ce9fe5a82b0375994 ... ffeb17d412dc2de0ac6e421f85eb15db
  ffef9246733b290086fd66e084905567
colData names(44): CellID analysis_protocol.protocol_core.protocol_id
  ... specimen_from_organism.provenance.document_id total_umis
rowGraphs(0): NULL
colGraphs(0): NULL
> range(colSums(assay(loom)[, 1:1000] > 0))
[1] 502 4660
```

- The size of the experiment, compared to full version retrieved
  earlier, and the range of genes per cell give us confidence that the
  filters have been applied correctly.

# 'Low-level' operations

```{r}
hca %>%
    tags() %>%
    dplyr::filter(tag == "v1") %>%
    dplyr::select(operation)

hca$matrix.lambdas.api.v1.core.post_matrix
```

- End-points exposed to the R user.
- `matrix.lambdas.api.v1.core.post_matrix()` requests construction of
  a customized count matrix, based on a series of filters, metadata
  fields, etc.

# Summary

- [HCAMatrixBrowser][] allows programmatic access to the HCA Matrix
  Service.
- Both complete and data filtered to precise criteria can be
  retrieved.
- The full RESTful HCA Matrix Service API is available for further
  customization.

## Session info

```{r}
sessionInfo()
```
