---
title: "Even more in-depth data discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Even more in-depth data discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE
  )
library(HCAExplorer)
```

Objectives:

- Use the [HCAExplorer][] package to identify fastq files for primary
  analysis.

[HCAExplorer]: https://bioconductor.org/packages/HCAExplorer

# The HCAExplorer package

```{r}
library(HCAExplorer)
print(citation("HCAExplorer"), style = "text")
```

Built on the [HCA azul][] RESTful API.

[HCA azul]: https://service.explore.data.humancellatlas.org 

```{r}
hca <- HCAExplorer()
hca
```

# Filtering to a specific project

Identify our favorite project and retrieve detailed metadata about it.

```{r}
title <- "Assessing the relevance of organoids to model inter-individual variation"
hca %>%
    filter(projectTitle %in% title) %>%
    results() %>%
    str()
```

# Obtaining a 'manifest' of fastq files.

```{r}
manifest <-
    hca %>%
    filter(projectTitle %in% title) %>%
    getManifest(fileFormat = "fastq.gz")
```

- Manifest created by the service.

```{r}
colnames(manifest)
```

- Considerable information available about each file.

```{r}
manifest %>%
    select(file_name, file_uuid) %>%
    distinct()
```

- File names mapped to file UUIDs.
- Files can be retrieved by UUID, e.g., using the HCA CLI (command
  line interface).

# Summary

- [HCAExplorer][] provides more flexible, but also more complicated,
  ways to explore the data
- Enables tasks like identifying and downloading all FASTQ files
  associated with a project.

## Session info

```{r}
sessionInfo()
```

